Abstractions for Type-Level Programming

### 2022-03-28 preprint.tex
Partially address Viktor's comments

During type normalization, our system evaluates pattern matching expressions according to Scala's runtime semantics.
Patterns are checked top-to-bottom, and type-tests are evaluated using runtime type information available after type erasure.

@@

\noindent
We could have equivalently defined |getName| by omitting the |dependent| qualifier and writing its result type  explicitly.

@@

Similarly to how we model other side effects, exceptions are approximated in types.
Our type-inference algorithm uses a new error type, |Error(e)|, which we infer when raising an exception with |throw e|.
Exception handlers are typed imprecisely using the default mode of type-inference.
Exceptions thrown in statement positions are not reflected in singleton types, since the type of |{e1; e2}| is simply |{ e2 }|.
However, exceptions thrown in tail positions (such as in remove from \Cref{sec:motivating-example}) can lead to types normalizing to |Error(e)|.
In these cases, our type system can prove that the program execution will encounter exceptional behavior, and report a compilation error.
This approach is conservative in that it might reject programs that recover from exceptions.
Also note that this is a sanity check, rather than a guarantee of no exceptions occurring at runtime.
That is, depending on which rules are used during subtyping, it is possible to succeed without entering type normalization, resulting in such errors going undetected.
Despite these shortcomings, our treatment of exceptions results in a practical way to raise compile-time errors.
It would be interesting to explore the addition of an effect system to our Scala extension.

>>>>

During type normalization, our system evaluates pattern matching expressions according to Scala's runtime semantics.
Patterns are checked top-to-bottom, and type-tests are evaluated using subtyping checks.

@@

\noindent
We could have equivalently defined |getName| by omitting the |dependent| qualifier and writing its result type explicitly.
The difference between the two would only be matter of syntax.

@@

Similarly to how we model other side effects, exceptions are approximated in types.
Our type-inference algorithm uses a new error type, |Error(e)|, which we infer when raising an exception with |throw e|.
Exception handlers are typed imprecisely using the default mode of type-inference.
Exceptions thrown in statement positions are not reflected in singleton types, since the type of |{ e1; e2 }| is simply |{ e2 }|.
However, exceptions thrown in tail positions (such as in remove from \Cref{sec:motivating-example}) can lead to types normalizing to |Error(e)|.
In these cases, our type system can prove that the program execution will encounter exceptional behavior, and report a compilation error.
This approach is conservative in that it might reject programs that recover from exceptions.
Also note that this is a sanity check, rather than a guarantee of no exceptions occurring at runtime.
That is, depending on which rules are used during subtyping, it is possible to succeed without entering type normalization, resulting in such errors going undetected.
Despite these shortcomings, our treatment of exceptions results in a practical way to raise compile-time errors.
It would be interesting to explore the addition of an effect system to our Scala extension.

### 2022-03-28 thesis.tex
Partially address Viktor's comments

\noindent
The |Remove| trait takes 3 type parameters: 2 inputs, |V| (the element to remove) and |Ps| (the list), and one output, |Out| (the list, with the element removed).
The two implicit definitions, |casehead| and |casetail|, correspond to the base case and the recursive case of the list removal operation, respectively.
The |Remove| trait is intended to be used as an implicit parameter, with constrained input types, and an unconstrained output type.

@@

\noindent
The |Singleton| type bound is a marker that instructs type inference to preserve literal singleton types (by default, the compiler widens those types).
The "|?|" type is Scala~3's new syntax for wildcard types~\citep[Wildcard Arguments in Types]{odersky2013scala}.

@@

>>>>

\noindent
The |Remove| trait takes 3 type parameters: 2 inputs, |V| (the element to remove) and |Ps| (the list), and one output, |Out| (the list, with the element removed).
The two implicit definitions, |casehead| and |casetail|, correspond to the base case and the recursive case of the list removal operation, respectively.
The right-hand side of those definitions is devoid of meaning, instances of the |Remove| trait merely act as placeholders (the type parameters of those constructor calls are inferred from their expected type).
The |Remove| trait is intended to be used as an implicit parameter, with constrained input types, and an unconstrained output type.

@@

\noindent
The |Singleton| type bound is a marker that instructs type inference to preserve union types and literal singleton types (by default, the compiler widens those types).
The "|?|" type is Scala~3's new syntax for wildcard types~\citep[Wildcard Arguments in Types]{odersky2013scala}.

@@

% TODO: Compare to singleton types
